---
title: "CAT Risk and Florida CAT Claims"
output: html_document
date: "2026-02-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning
# Load libraries
```{r, message = FALSE}
library(tidyverse)
library(dplyr)
library(stringr)
library(readxl)
library(ggplot2)
library(fitdistrplus)
library(tidycensus)
library(zipcodeR)
```

# Read and Load data

The HURDAT2 dataset (21 columns, 55230 rows) provides storm position and intensity at six-hour intervals. While the data includes detailed wind-field information (column 9-21), this study focuses on storm frequency and maximum sustained wind speed, which are sufficient proxies for hurricane severity in an insurance loss modeling context.

NOTE: This is the base hurricane storm set; essential to get annual frequencies used in hurricane model validation.

```{r, message = FALSE}
# url <- "https://www.nhc.noaa.gov/data/hurdat/hurdat2-1851-2024-040425.txt"
lines <- read_excel("HURDAT2.xlsx", col_names = FALSE)
```

## Explaining columns
1. id = Basin (AL) + Storm # (04) + Year (1980)
2. Status:
TD = Tropical Depression,
TS = Tropical Storm,
HU = Hurricane,
EX = Extratropical Cyclone,
LO = Low pressure system,
DB = Disturbance,
SS = Subtropical Storm,
SD = Subtropical Depression,
WV = Tropical Wave
3. record_id: landfall flag is in here

# Clean Data
```{r, message = FALSE}
dt <- list()
storm_name <- NA
storm_id <-NA
for (i in 1:nrow(lines)) {
  l <- lines[i, ]
  if (str_detect(l, "^\\s*AL")) {
    storm <- str_split(l, ",")[[1]]
    storm_id <- str_trim(storm[1])
    storm_name <- str_trim(storm[2])
  } else {
    storm <- str_split(l, ",")[[1]]
    dt[[length(dt) + 1]] <- data.frame(
      id         = storm_id,
      name       = storm_name,
      date       = str_trim(storm[1]),
      time       = str_trim(storm[2]),
      record_id  = str_trim(storm[3]),
      status     = str_trim(storm[4]),
      lat        = str_trim(storm[5]),
      long       = str_trim(storm[6]),
      wind       = as.numeric(str_trim(storm[7])),
      pressure   = as.numeric(str_trim(storm[8])),
      
      # 34-kt wind radii
      ne34       = as.numeric(str_trim(storm[9])),
      se34       = as.numeric(str_trim(storm[10])),
      sw34       = as.numeric(str_trim(storm[11])),
      nw34       = as.numeric(str_trim(storm[12])),
      
      # 50-kt wind radii
      ne50       = as.numeric(str_trim(storm[13])),
      se50       = as.numeric(str_trim(storm[14])),
      sw50       = as.numeric(str_trim(storm[15])),
      nw50       = as.numeric(str_trim(storm[16])),
      
      # 64-kt wind radii
      ne64       = as.numeric(str_trim(storm[17])),
      se64       = as.numeric(str_trim(storm[18])),
      sw64       = as.numeric(str_trim(storm[19])),
      nw64       = as.numeric(str_trim(storm[20])),
      
      # Radius of max wind
      r_max_wind = as.numeric(str_trim(storm[21]))
    )
  }
}

hurdat <- bind_rows(dt)
```

```{r, message = FALSE}
tail(hurdat)
```


Due to advances in aircraft reconnaissance, data after 1944 is considered more reliable (?). Also keeping Florida data.
```{r, message = FALSE}
hurdat_cl <- hurdat %>% 
  mutate(status = str_trim(status),
    date = ymd(date),
    year = year(date),
    month = month(date),
    day = day(date),
    lat = case_when(str_ends(str_trim(lat),  "N") ~  as.numeric(str_remove(str_trim(lat),  ".$")),
      str_ends(str_trim(lat),  "S") ~ -as.numeric(str_remove(str_trim(lat),  ".$"))),
    long = case_when(str_ends(str_trim(long), "W") ~ -as.numeric(str_remove(str_trim(long), ".$")),
      str_ends(str_trim(long), "E") ~  as.numeric(str_remove(str_trim(long), ".$"))),
    wind_mph = wind * 1.15078,
    pressure = na_if(pressure, -999),
    pressure = na_if(pressure, 0),
    across(ne34:r_max_wind, ~ na_if(.x, -999)))  %>% 
  filter(year >= 1900,
         status %in% c("HU", "TS", "SS"))
```

```{r, message = FALSE}
head(hurdat_cl)
```

```{r, message = FALSE}
storm_summary <- hurdat_cl %>%
  group_by(id, name, year) %>%
  summarise(
    peak_wind_mph = max(wind_mph, na.rm = TRUE),
    min_pressure = min(pressure, na.rm = TRUE),
    peak_month = month[which.max(wind_mph)],
    peak_status = case_when(
      any(status == "HU") ~ "HU",
      any(status == "SS") ~ "SS",
      TRUE ~ "TS"
    ),
    landfall = any(record_id=="L"),
    mean_r34 = mean(c(ne34, se34, sw34, nw34), na.rm = TRUE),
    mean_r64 = mean(c(ne64, se64, sw64, nw64), na.rm = TRUE),
    r_max_wind = mean(r_max_wind, na.rm = TRUE),
    .groups = "drop"
  )
```

Filter out Florida storms
```{r, message = FALSE}
florida_ids <- hurdat_cl %>%
  filter(lat  >= 23.0 & lat  <= 32.0,
         long >= -88.0 & long <= -79.0) %>%
  distinct(id)

florida_summary <- storm_summary %>%
  semi_join(florida_ids, by = "id")
```



# EDA
## Wind
Saffir-Simpson Hurricane Wind Scale (1):

Category    Winds (mph)   Damage
1           74 – 95       Minimal
2           96 – 110      Moderate
3           111 – 129     Extensive
4           130 – 156     Extreme
5           >157          Catastrophic
```{r, echo=FALSE, message=FALSE}
# Catergorize by Saffir-Simpson Scale
florida_summary <- florida_summary %>%
  mutate(
    category = case_when(
      peak_wind_mph >= 157 ~ "Cat 5",
      peak_wind_mph >= 130 & peak_wind_mph < 157 ~ "Cat 4",
      peak_wind_mph >= 111 & peak_wind_mph < 130 ~ "Cat 3",
      peak_wind_mph >= 96  & peak_wind_mph < 111 ~ "Cat 2",
      peak_wind_mph >= 74  & peak_wind_mph < 96 ~ "Cat 1",
      peak_wind_mph >= 39  & peak_wind_mph < 74 ~ "TS",
      TRUE ~ "TD"
    ),
    # Order the factor for plotting
    category = factor(category, 
                      levels = c("TD", "TS", "Cat 1", "Cat 2", 
                                 "Cat 3", "Cat 4", "Cat 5"))
  )


storm_summary <- storm_summary %>%
  mutate(
    category = case_when(
      peak_wind_mph >= 157 ~ "Cat 5",
      peak_wind_mph >= 130 & peak_wind_mph < 157 ~ "Cat 4",
      peak_wind_mph >= 111 & peak_wind_mph < 130 ~ "Cat 3",
      peak_wind_mph >= 96  & peak_wind_mph < 111 ~ "Cat 2",
      peak_wind_mph >= 74  & peak_wind_mph < 96 ~ "Cat 1",
      peak_wind_mph >= 39  & peak_wind_mph < 74 ~ "TS",
      TRUE ~ "TD"
    ),
    # Order the factor for plotting
    category = factor(category, 
                      levels = c("TD", "TS", "Cat 1", "Cat 2", 
                                 "Cat 3", "Cat 4", "Cat 5"))
  )
```

```{r, echo=FALSE, message=FALSE}
# Category distribution
florida_summary %>%
  count(category) %>%
  ggplot(aes(x = category, y = n, fill = category)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(
    "TD"    = "gray",
    "TS"    = "steelblue",
    "Cat 1" = "yellow",
    "Cat 2" = "orange",
    "Cat 3" = "darkorange",
    "Cat 4" = "red",
    "Cat 5" = "darkred"
  )) +
  labs(title = "Florida Storms by Saffir-Simpson Category")


storm_summary %>%
  count(category) %>%
  ggplot(aes(x = category, y = n, fill = category)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(
    "TD"    = "gray",
    "TS"    = "steelblue",
    "Cat 1" = "yellow",
    "Cat 2" = "orange",
    "Cat 3" = "darkorange",
    "Cat 4" = "red",
    "Cat 5" = "darkred"
  )) +
  labs(title = "Atlantic Storms by Saffir-Simpson Category")
```
```{r}
library(diptest)
dip.test(storm_summary$peak_wind_mph)
```


```{r, echo=FALSE, message=FALSE}
# Peak wind distributions
ggplot(florida_summary, aes(x = peak_wind_mph, fill = peak_status)) +
  geom_bar(position = "stack") +
  labs(title = "Peak Wind Distribution of Florida storms")
```

## Storm count
Overall number storms by status
```{r, echo=FALSE, message=FALSE}
florida_summary %>% count(peak_status)
```

Overall number of landfall
```{r, echo=FALSE, message=FALSE}
florida_summary %>%  count(landfall)
```

```{r, include=FALSE, message=FALSE}
annual_counts_at <- storm_summary %>% count(year)
m_at <- mean(annual_counts_at$n)
v_at <- var(annual_counts_at$n)
v_at/m_at # Overdispersion of Atlantic storm count


annual_counts_fl <- florida_summary %>% count(year)
m_fl <- mean(annual_counts_fl$n)
v_fl <- var(annual_counts_fl$n)
v_fl/m_fl # Overdispersion of Florida storm count


annual_counts_lf <- florida_summary %>% filter(landfall == TRUE) %>% count(year)
m_lf <- mean(annual_counts_lf$n)
v_lf <- var(annual_counts_lf$n)
v_lf/m_lf # Overdispersion of Florida landfall count
```

Plots on storm count per year (red line is the avarage)
```{r, echo=FALSE, message=FALSE}
storm_summary %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_bar(stat = "identity") +
  geom_line(aes(y = m_at, colour = "red")) +
  labs(title = "Storm count per year in Atlantic Basin")
```

```{r}
florida_summary %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_bar(stat = "identity") +
  geom_line(aes(y = m_fl, colour = "red")) +
  labs(title = "Storm count per year in Florida")
```
"The Poisson assumption is supported by historical data. The Poisson rates are in good agreement with historical data" (1)

Poisson is a good choice since Florida's overdispersion is near 1 (i.e. mean = variance). Also, looking at the distribution, number of storms varies randomly, and knowing about past storm count says nothing about future storm count (memoryless property).

NOTE: if overdispersion >> 1 (i.e. some years has more storms than others; fat tail), choose Neg Bin


## Landfall
```{r}
florida_summary %>% 
  filter(landfall == TRUE) %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_bar(stat = "identity") +
  geom_line(aes(y = m_lf, colour = "red")) +
  labs(title = "Landfall per year in Florida", color = "Mean")
```
```{r}
landfall_region <- hurdat_cl %>% 
  filter(record_id == "L") %>% 
  mutate(region = case_when(
    lat >= 29.5 ~ "Panhandle",
    lat >= 27.5 & lat < 29.5 ~ "North Florida",
    lat >= 25.5 & lat < 27.5 ~ "Central Florida",
    lat <  25.5 ~ "South Florida"
  ))

landfall_region %>%  count(region, year) %>% 
  ggplot(aes(x = year, y = n, colour = region)) +
  geom_line()
```
South Florida > Panhandle > North Florida > Central Florida

```{r}
# How severe?
florida_summary %>% 
  left_join(landfall_region %>% 
              dplyr::select(id, region), by = "id") %>% 
  group_by(region) %>% 
  summarise(n = n(),
            mean_wind = mean(peak_wind_mph),
            major_storm = mean(category %in% c("Cat 3","Cat 4","Cat 5")) * 100) #percentage of major storms
```

```{r}
# Forward Speed + Track direction

landfall_details <- hurdat_cl %>%
  filter(record_id == "L",
         lat >= 23.0 & lat <= 32.0,
         long >= -88.0 & long <= -79.0) %>%
  group_by(id) %>%
  slice(1) %>%
  dplyr::select(id, lat, long, wind_mph, pressure, r_max_wind)

head(landfall_details)
```


## Seasonality
```{r, echo=FALSE, message=FALSE}
florida_summary %>% 
  count(peak_month, peak_status) %>% 
  ggplot(aes(x=peak_month, y = n, fill = peak_status)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Florida Storm Count by Month and Status")
```

```{r, echo=FALSE, message=FALSE}
# Category by month
florida_summary %>%
  count(peak_month, category) %>%
  ggplot(aes(x = peak_month, y = n, fill = category)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_x_continuous(breaks = 1:12) +
  scale_fill_manual(values = c(
    "TD"    = "gray",
    "TS"    = "steelblue",
    "Cat 1" = "yellow",
    "Cat 2" = "orange",
    "Cat 3" = "darkorange",
    "Cat 4" = "red",
    "Cat 5" = "darkred"
  )) +
  labs(title = "Storm Severity by Month")
```

How it has changed through years?

```{r, echo=FALSE, message=FALSE}
florida_summary %>% 
  mutate(era = case_when(year < 1950 ~ "1900-1949",
                        year < 2000 ~ "1950-1999",
                        TRUE ~ "2000-present")) %>% 
  count(era, peak_month) %>% 
  group_by(era) %>% 
  mutate(storm_count = n/sum(n)) %>% 
  ggplot(aes(x=peak_month, y=storm_count, fill=era)) +
  geom_bar(stat = "identity", position = "dodge") + 
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Seasonal Storm Count by Era")
```

```{r, echo=FALSE, message=FALSE}
florida_summary %>%
  mutate(era = case_when(
    year < 1950 ~ "1900-1949",
    year < 2000 ~ "1950-1999",
    TRUE        ~ "2000-present"
  )) %>%
  count(era, category) %>%
  group_by(era) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = category, y = prop, fill = category)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ era) +
  scale_fill_manual(values = c(
    "TS"    = "steelblue",
    "Cat 1" = "yellow",
    "Cat 2" = "orange",
    "Cat 3" = "darkorange",
    "Cat 4" = "red",
    "Cat 5" = "darkred"
  )) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Storm Category Proportion by Era",
       x = "Category", y = "Proportion")
```

## Overall
- Overall, TS has highest frequency, following by CAT 1, CAT 4, CAT 2. This doesn't change through eras. --> a mixture distribution for severity
- It is clearly that storms start in 5 and peak at Month 9 in all eras. Month 8-10 have the highest frequency and severity (including CAT 5). --> conditional on peak season
- In the span of 25 years recently, there are more CAT 5 storms happened proportionally. --> climate trend?

```{r}
ggplot(florida_summary, aes(x = peak_wind_mph)) +
  geom_density(fill = "steelblue", alpha = 0.5, linewidth = 1) +
  geom_vline(xintercept = c(74, 96, 111, 130, 157),
             linetype = "dashed", color = "red", alpha = 0.7) +
  annotate("text", x = c(85, 103, 120, 143, 170), y = 0.002,
           label = c("Cat1", "Cat2", "Cat3", "Cat4", "Cat5"),
           color = "red", size = 3) +
  labs(title = "Wind Speed Density",
       x = "Peak Wind Speed (mph)", y = "Density")
```

### Fit mixture distribution
fits a mixture of normal distributions using the EM (Expectation-Maximization) algorithm
k = 2 — find 2 components

The algorithm iterates back and forth between:
E step — assign each storm a probability of belonging to component 1 or 2
M step — update the parameters (mean, sd, weight) to best fit those assignments
Repeats until it converges on stable parameters

```{r}
library(mixtools)

mix_fit <- normalmixEM(florida_summary$peak_wind_mph, k = 2)
summary(mix_fit)

# Extract parameters cleanly
tibble(
  component = c("Weak (TS/Cat1/Cat2)", "Intense (Cat3-5)"),
  lambda    = mix_fit$lambda,   # mixing weights
  mu        = mix_fit$mu,       # mean wind speed
  sigma     = mix_fit$sigma     # standard deviation
)
```


# Monte Carlo
```{r, message = FALSE}
# Get Poisson lambda
fit_poisson <- fitdist(annual_counts_fl$n, "pois")
lambda_annual <- fit_poisson$estimate["lambda"]
```
## Frequency
```{r}
set.seed(100)
n_sim <- 100000

# Number of storms per year
annual_storm <- rpois(n_sim, lambda = lambda_annual)

table(annual_storm) %>% head(10)
```

## Severity
```{r}
# Wind speed
total_storm <- sum(annual_storm)

component <- sample(
  x       = 1:2,
  size    = total_storm,
  replace = TRUE,
  prob    = mix_fit$lambda  # c(0.485, 0.515)
)

# Step 2b — draw wind speed from that component's normal distribution
wind_speeds <- rnorm(
  n    = total_storm,
  mean = mix_fit$mu[component],     # 62 or 120 mph
  sd   = mix_fit$sigma[component]   # 13 or 32 mph
)

# Step 2c — floor at 39 mph (minimum TS threshold)
wind_speeds <- pmax(wind_speeds, 39)

# Sanity checks
mean(wind_speeds)
quantile(wind_speeds, c(0.25, 0.5, 0.75, 0.90, 0.99))
sum(wind_speeds >= 157) / total_storm  # % Cat 5 storms
```

## Vulnerability
```{r}
vulnerability <- function(wind_mph, construction = "masonry") {
  params <- list(
    manufactured = list(center = 80,  steepness = 0.12),
    wood_frame   = list(center = 95,  steepness = 0.10),
    masonry      = list(center = 110, steepness = 0.08)
  )
  p <- params[[construction]]
  1 / (1 + exp(-p$steepness * (wind_mph - p$center)))
}
```


```{r, eval=FALSE}
census_api_key("7adee0ef167edfd313a5a87ef9a3287b3c547249", install = TRUE, overwrite = TRUE)
```

```{r}
readRenviron("~/.Renviron")

fl_exposure <- get_acs(
  geography = "zcta",          # ZIP Code Tabulation Area
  variables = c(
    housing_units = "B25001_001",   # total housing units
    median_value  = "B25077_001"    # median home value ($)
  ),
  year   = 2022,
  output = "wide"
) %>%
  # filter to Florida ZIPs (start with 32 or 33 or 34)
  filter(str_starts(GEOID, "32") |
         str_starts(GEOID, "33") |
         str_starts(GEOID, "34")) %>% 
  transmute(
    zip           = GEOID,
    housing_units = housing_unitsE,
    median_value  = median_valueE,
    TIV           = housing_units * median_value
  ) %>%
  filter(!is.na(TIV), TIV > 0)

nrow(fl_exposure)          # how many Florida ZIPs?
sum(fl_exposure$TIV) / 1e9 # total insured value in billions
fl_exposure %>% arrange(desc(TIV)) %>% head(10)  # top 10 ZIPs
```

```{r}
zip_coords <- zip_code_db %>%
  filter(state == "FL") %>%
  dplyr::select(zipcode, lat, lng) %>%
  rename(zip = zipcode, zip_lat = lat, zip_long = lng)

fl_exposure <- fl_exposure %>%
  left_join(zip_coords, by = "zip") %>%
  filter(!is.na(zip_lat), !is.na(zip_long))

# Plot
ggplot(fl_exposure, aes(x = zip_long, y = zip_lat, color = TIV / 1e6)) +
  geom_point(size = 0.8) +
  scale_color_viridis_c(name = "TIV ($M)", trans = "log10") +
  labs(title = "Florida ZIP Code Exposure (TIV)",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

```{r}

```




# NOTES
## Distributions
exp: (Exponential) The parameter is rate (defaults to 1). The mean of the distribution is 1/rate.
norm: (Normal)The most famous distribution in statistics, this is the well-known “bell-curve”. Parameters of the distribution are mean (defaults to 0) and sd (defaults to 1).
unif: (Uniform) Parameters are min (defaults to 0) and max (defaults to 1).
binom: (Binomial) Parameters are size, the number of trials, and prob, the probability of success on a single trial (no defaults).
geom: (Geometric) Parameter is prob, the probability of success on each independent trial. Note that the distribution is defined in terms of the number of failures before the first success.
pois: (Poisson) The parameter is lambda, the mean.

## r code chunk notation
echo: display code in output?
eval: run code?
include: display code and results?
message: display warning message?



# Source
(1) https://fchlpm.sbafla.com/media/4fdg1eyk/rms21standardsdisclosures_05192023.pdf
